---
title: "DAS and doubleDAS User Guide"
format: pdf
editor: visual
vignette: >
  %\VignetteIndexEntry{DAS and doubleDAS User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# DAS

## DAS Parameters

The following DAS parameters define the population, the number of sample points to draw from the population and the maximum number of candidate samples.

```{r}
set.seed(2026)

# Numeric matrix of population coordinates, one unit per row.
# x points uniformly in [0,1)^2
myPop <- matrix(runif(2000 * 2), ncol = 2)

# The number of sample points to draw from the population.
n <- 250

# Maximum number of candidate samples, \code{2 <= J1 <= floor(N/2)}.
# Default is \code{min(200, floor(N/2))}. Values above 500 are allowed but
# trigger a warning (large assignment problems dominate run time).
J1 <- 200
```

### Additional DAS Parameters

The following are additional parameters supported by DAS.

- n_threads
  - Threads used when filling the cost matrix. A value of $0$ uses every logical CPU. Parallel fill runs only when the current $J$ is at least $256$ and $n\_threads$ is greater than $1$. Default is $1$.
- verbose
  - Print step timings. Default is FALSE.
- cache_W
  - If $TRUE$, build one $N \times N$ inverse-distance matrix in C++. If $FALSE$, compute inverse distances on the fly. Default is $TRUE$ when $N <= 8000$, otherwise $FALSE$.

```{r}
# Use all logical CPU's
n_threads <- 0

# Display step timings
verbose <- TRUE

# Build single inverse-distance matrix
cache_W <- TRUE
```

## Generate Sample

Call DAS to generate the sample.

```{r}
library(spbalDAS)

sampleMatrix <- DAS(pop = myPop,
                    n = n, 
                    J1 = J1, 
                    n_threads = n_threads, 
                    verbose = verbose,
                    cache_W = cache_W)
```

Show first candidate sample.

```{r}
# first candidate sample
s <- sampleMatrix[1,]
s
```

Plot the sample.

```{R}
library(reshape2)
library(ggplot2)

# plotting...
index_matrix <- sampleMatrix # leave sampleMatrix intact
df <- myPop

# Melt the matrix into long format to map indices to colors
index_long <- melt(index_matrix)
colnames(index_long) <- c("row", "col", "index")

# Merge the matrix data with the dataframe coordinates
plot_data <- merge(index_long, df, by.x = "index", by.y = "row.names")

# Create a color column for the row groups of the matrix
plot_data$color_group <- ifelse(plot_data$row == 1, "red", "black")

# Dynamic title using the actual values of n and number of locations
n_value <- n
n_locations <- length(myPop) / 2

# Plot the points - BLACK drawn first, RED on top (guaranteed layering)
ggplot(plot_data, aes(x = V1, y = V2, color = color_group)) +
  # Black points (bottom layer)
  geom_point(data = plot_data[plot_data$color_group == "black", ],
             size = 3) +
  # Red points (top layer - drawn after black)
  geom_point(data = plot_data[plot_data$color_group == "red", ],
             size = 3) +
  scale_color_manual(values = c("black" = "black", "red" = "red")) +
  theme_minimal() +
  labs(color = "Matrix Row Group",
       title = paste0("DAS Sample of n = ", n_value,
                      " from ", n_locations, " Locations, J1=", J1),
       caption = "First row colored in RED.") +
  xlab("X Coordinate") +
  ylab("Y Coordinate")
```

# doubleDAS

## doubleDAS Parameters

The following doubleDAS parameters define the population, the auxiliary varaibles, the number of sample points to draw from the population, the mixing parameter and the maximum number of candidate samples.

```{r}
set.seed(2026)

# Numeric matrix of population coordinates, one unit per row.
# x points uniformly in [0,1)^2
myPop <- matrix(runif(2000 * 2), ncol = 2)

# Numeric matrix of auxiliary variables, one row per unit.
myAux <- matrix(runif(2000 * 2), ncol = 2)
myAux <- myAux / sqrt(rowSums(myAux^2)) 

# The number of sample points to draw from the population.
n <- 150

# Mix in [0, 1]. Default 0.5.
alpha <- 0.5

# Maximum number of candidate samples, \code{2 <= J1 <= floor(N/2)}.
# Default is \code{min(200, floor(N/2))}. Values above 500 are allowed but
# trigger a warning (large assignment problems dominate run time).
J1 <- 200
```

### Additional doubleDAS Parameters

The following are additional parameters supported by DAS.

- n_threads
  - Threads used when filling the cost matrix. A value of $0$ uses every logical CPU. Parallel fill runs only when the current $J$ is at least $256$ and $n\_threads$ is greater than $1$. Default is $1$.
- verbose
  - Print step timings. Default is FALSE.
- cache_W
  - If $TRUE$, build one $N \times N$ inverse-distance matrix in C++ when $alpha > 0$. If $FALSE$, compute inverse distances on the fly. Default is $TRUE$ when $N <= 2500$, otherwise $FALSE$. At large $N$ this is usually slower than on-the-fly distances.

```{r}
# Use all logical CPU's
n_threads <- 0

# Display step timings
verbose <- TRUE

# Build single inverse-distance matrix
cache_W <- TRUE
```

## Generate Sample

Call doubleDAS to generate the sample.

```{r}
library(spbalDAS)

sampleMatrix <- doubleDAS(pop = myPop,
                          aux = myAux,
                          n = n,
                          alpha = alpha,
                          J1 = J1, 
                          n_threads = n_threads, 
                          verbose = verbose,
                          cache_W = cache_W)
```

Show first candidate sample.

```{r}
# first candidate sample
s <- sampleMatrix[1,]
s
```

Plot the sample.

```{R}
library(reshape2)
library(ggplot2)

# plotting...
index_matrix <- sampleMatrix # leave sampleMatrix intact
df <- myPop

# Melt the matrix into long format to map indices to colors
index_long <- melt(index_matrix)
colnames(index_long) <- c("row", "col", "index")

# Merge the matrix data with the dataframe coordinates
plot_data <- merge(index_long, df, by.x = "index", by.y = "row.names")

# Create a color column for the row groups of the matrix
plot_data$color_group <- ifelse(plot_data$row == 1, "red", "black")

# Dynamic title using the actual values of n and number of locations
n_value <- n
n_locations <- length(myPop) / 2

# Plot the points, colored by the row group of the matrix
ggplot(plot_data, aes(x = V1, y = V2, color = color_group)) +
  # Black points (bottom layer)
  geom_point(data = plot_data[plot_data$color_group == "black", ],
             size = 3) +
  # Red points (top layer - drawn after black)
  geom_point(data = plot_data[plot_data$color_group == "red", ],
             size = 3) +
  scale_color_manual(values = c("black" = "black", "red" = "red")) +
  theme_minimal() +
  labs(color = "Matrix Row Group",
       title = paste0("doubleDAS Sample of n = ", n_value,
                      " from ", n_locations, " Locations, J1=", J1),
       subtitle = paste0("alpha=", alpha, "."),
       caption = "First row colored in RED.") +
  xlab("X Coordinate") +
  ylab("Y Coordinate")
```

## The End
